---
title: "Housing Supply Elasticity"
subtitle: "Per-state OLS, 2010–2024: level–level and log–log"
format:
  html:
    embed-resources: true
    toc: true
    fig-width: 12
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(arrow)
library(dplyr)
library(tidyr)
library(purrr)
library(forcats)
library(ggplot2)
library(scales)
library(tigris)
library(sf)
library(patchwork)
library(ggrepel)

options(tigris_use_cache = TRUE)

panel <- read_parquet("data/panel.parquet")
```

```{r fit}
elasticity_linear <- panel |>
  filter(!is.na(permits_per_1000_pop), !is.na(hpi_at_annual)) |>
  group_by(fips_state, state_name, state_abbr) |>
  filter(n() >= 3) |>
  nest() |>
  mutate(
    slope = map_dbl(data, \(d)
      coef(lm(permits_per_1000_pop ~ hpi_at_annual, data = d))[["hpi_at_annual"]]
    )
  ) |>
  select(!data) |>
  ungroup() |>
  mutate(slope_norm = pmax(0, slope) / max(pmax(0, slope), na.rm = TRUE))

elasticity_loglog <- panel |>
  filter(!is.na(permits_per_1000_pop), !is.na(hpi_at_annual),
         permits_per_1000_pop > 0, hpi_at_annual > 0) |>
  group_by(fips_state, state_name, state_abbr) |>
  filter(n() >= 3) |>
  nest() |>
  mutate(
    coef = map_dbl(data, \(d)
      coef(lm(log(permits_per_1000_pop) ~ log(hpi_at_annual), data = d))[["log(hpi_at_annual)"]]
    )
  ) |>
  select(!data) |>
  ungroup() |>
  mutate(coef_norm = pmax(0, coef) / max(pmax(0, coef), na.rm = TRUE))
```

## Raw Data: Level–Level and Log–Log

One dot per state-year. The line is a single OLS fit across all observations;
per-state slopes (computed above) are the basis for the choropleth and bar charts.

```{r raw-scatter}
scatter_data <- panel |>
  filter(!is.na(permits_per_1000_pop), !is.na(hpi_at_annual),
         permits_per_1000_pop > 0, hpi_at_annual > 0)

p_raw_lin <- ggplot(scatter_data, aes(x = hpi_at_annual, y = permits_per_1000_pop)) +
  geom_point(alpha = 0.25, size = 1.2, color = "#888888") +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              color = "#d6604d", fill = "#d6604d", alpha = 0.15, linewidth = 0.9) +
  scale_x_continuous(labels = comma) +
  labs(
    x = "HPI (all-transactions)",
    y = "Permits per 1,000 pop",
    title = "Level–Level"
  ) +
  theme_minimal(base_size = 11)

p_raw_log <- ggplot(scatter_data, aes(x = log(hpi_at_annual), y = log(permits_per_1000_pop))) +
  geom_point(alpha = 0.25, size = 1.2, color = "#888888") +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              color = "#d6604d", fill = "#d6604d", alpha = 0.15, linewidth = 0.9) +
  labs(
    x = "log(HPI)",
    y = "log(Permits per 1,000 pop)",
    title = "Log–Log"
  ) +
  theme_minimal(base_size = 11)

p_raw_lin + p_raw_log
```

```{r geo}
states_sf <- states(cb = TRUE, resolution = "20m", year = 2022) |>
  filter(!STUSPS %in% c("PR", "GU", "VI", "MP", "AS")) |>
  shift_geometry()

lin_sf <- states_sf |> left_join(elasticity_linear, by = c("GEOID" = "fips_state"))
ll_sf  <- states_sf |> left_join(elasticity_loglog,  by = c("GEOID" = "fips_state"))

# Normalized position of unit elasticity (β=1) on the [0,1] scale
ll_ref_norm <- 1 / max(pmax(0, elasticity_loglog$coef), na.rm = TRUE)
```

## Choropleths

```{r choropleths}
p_choro_lin <- ggplot(lin_sf) +
  geom_sf(aes(fill = slope_norm), color = "white", linewidth = 0.3) +
  scale_fill_gradient(
    low = "grey92", high = "#d6604d", limits = c(0, 1),
    name = "Normalized slope"
  ) +
  labs(title = "Linear: Level–Level OLS Slope") +
  theme_void(base_size = 11) +
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5))

p_choro_ll <- ggplot(ll_sf) +
  geom_sf(aes(fill = coef_norm), color = "white", linewidth = 0.3) +
  scale_fill_gradient(
    low = "grey92", high = "#d6604d", limits = c(0, 1),
    name = "Normalized β"
  ) +
  labs(title = "Log-Log: Elasticity Coefficient  (β = 1 → unit elasticity)") +
  theme_void(base_size = 11) +
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5))

p_choro_lin + p_choro_ll
```

## Ranked Bar Charts

```{r bars, fig.height=10}
p_bar_lin <- elasticity_linear |>
  mutate(state_abbr = fct_reorder(state_abbr, slope_norm)) |>
  ggplot(aes(x = slope_norm, y = state_abbr, fill = slope_norm)) +
  geom_col() +
  scale_fill_gradient(low = "grey92", high = "#d6604d", limits = c(0, 1), guide = "none") +
  labs(x = "Normalized slope", y = NULL, title = "Linear: Level–Level") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(hjust = 0.5))

p_bar_ll <- elasticity_loglog |>
  mutate(state_abbr = fct_reorder(state_abbr, coef_norm)) |>
  ggplot(aes(x = coef_norm, y = state_abbr, fill = coef_norm)) +
  geom_col() +
  geom_vline(xintercept = ll_ref_norm, linetype = "dashed", color = "gray40", linewidth = 0.5) +
  scale_fill_gradient(low = "grey92", high = "#d6604d", limits = c(0, 1), guide = "none") +
  labs(x = "Normalized β  (dashed = unit elasticity)", y = NULL,
       title = "Log-Log: Elasticity Coefficient") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(hjust = 0.5))

p_bar_lin + p_bar_ll
```

## Scatter: Elasticity vs. Average Supply

Do states with higher log-log elasticity actually permit more housing? The dashed
line marks unit elasticity (β = 1): supply growing proportionally with prices.

```{r scatter, fig.width=9, fig.height=7}
avg_permits <- panel |>
  filter(!is.na(permits_per_1000_pop)) |>
  group_by(fips_state, state_abbr, state_name) |>
  summarise(avg_permits = mean(permits_per_1000_pop), .groups = "drop")

elasticity_loglog |>
  left_join(avg_permits, by = c("fips_state", "state_abbr", "state_name")) |>
  ggplot(aes(x = coef_norm, y = avg_permits)) +
  geom_vline(xintercept = ll_ref_norm, linetype = "dashed", color = "gray50") +
  geom_point(aes(color = coef_norm), size = 3) +
  geom_text_repel(aes(label = state_abbr), size = 3, max.overlaps = 40) +
  scale_color_gradient(low = "grey70", high = "#d6604d", limits = c(0, 1), guide = "none") +
  labs(
    x = "Normalized Log-Log Elasticity β",
    y = "Avg Permits per 1,000 Pop (2010–2024)",
    title = "Log-Log Elasticity vs. Average Supply",
    subtitle = "Dashed line = unit elasticity (β = 1)"
  ) +
  theme_minimal(base_size = 12)
```
